---
title: "ST 558 HW 4 - Ben's version"
format: html
editor_options: 
  chunk_output_type: console
---

## Ben's work goes here

# Task 1 - Querying an API and a Tidy-Style Function

Return information on recent Apple news 

```{r}
library(httr)
library(tidyverse)
API_data <- GET("https://newsapi.org/v2/everything?q=apple&from=2026-06-10&to=2026-06-10&sortBy=popularity&apiKey=0009f78a3db94bc0b36a3a84d1cef3a7")
str(API_data)
```

Find the article information

```{r}
library(jsonlite)

# parse the url to find the content section
parsed <- fromJSON(rawToChar(API_data$content))
names(parsed)

#make it a tibble
articles <- parsed$articles |>
            as_tibble()
articles
```
Build a function to do this for us

```{r}
news_search <- function(search, start_date, api_key) {
  # get url
  response <- GET(url = "https://newsapi.org/v2/everything",
                  query = list(q = search,
                               from = start_date,
                               language = "en",
                               pageSize = 50,
                               apiKey = api_key))
  
  # find article information
  parsed <- fromJSON(rawToChar(response$content))
  
  # make it a tibble
  articles <- parsed$articles |>
    as_tibble()

  return(articles)
}
```

Let's test our function

```{r}
# 1. Given example
news_search("Iran", 2026-05-02, "0009f78a3db94bc0b36a3a84d1cef3a7")

# 2. Top business headlines
news_search("business", 2026-06-11, "0009f78a3db94bc0b36a3a84d1cef3a7")
```

Works as expected


# Task 2 - Summarizing Roller Coaster Data

Lets read in our data

```{r}
rcs_raw <- read.csv("https://www4.stat.ncsu.edu/online/datasets/635USrollercoasters.csv")

# remove some columns
rcs <- rcs_raw |>
  select(-c(Boundary, Latitude, Longitude))

# look at types
str(rcs)
```

Lets clean our data

```{r}
rcs <- rcs_raw |>
       select(-c(Boundary, Latitude, Longitude)) |> #remove some columns
       rename(Inversions =`Inversions.`, #rename some columns
              Inversions.Count = `X..of.Inversions`) |>
       mutate(across(c(Census.Region, AgeGroup, Scale, Type, Design, Inversions), 
                          ~ as.factor(.))) |> #turn some variables into factors
       as_tibble()

str(rcs)
```

## Basic EDA

Lets check the `summary` of the data to make sure everything looks okay

```{r}
summary(rcs)
```

Seems like everything checks out. Lets check the counts of missing values in the data by column

```{r}
colSums(is.na(rcs))
```

`Drop`, `TopSeed`, and `Duration` are the variables with the most missing values. They're missing 128, 74, and 66 values respectively. 


## Categorical Variables

Lets create some contingency tables using `table()`

```{r}
#one-way table
table(rcs$Scale, useNA = 'always')
```

From the table we can see there are `r table(rcs$Scale, useNA = 'always')[1]` extreme roller coasters in the dataset 

```{r}
#two-way table
table(rcs$Scale, rcs$Census.Region, useNA = 'always')
```

From the table we can see there are `r table(rcs$Scale, rcs$Census.Region, useNA = 'always')[1,1]` extreme roller coasters in the Midwest 

```{r}
#three-way table
table(rcs$Scale, rcs$Census.Region, rcs$AgeGroup, useNA = 'always')
```

From the table we can see there are `r table(rcs$Scale, rcs$Census.Region, rcs$AgeGroup, useNA = 'always')[1, 1, "1:Older"]` extreme roller coasters in the Midwest that are considered to be 'older'


Now lets create some conditional two way tables. We'll do this in two different ways:

- Once, by subsetting the data with `filter()` and then creating the two-way table

- Once, by creating a three-way table and subsetting it

```{r}
#subsetting with filter()
new_rcs <- rcs |>
  filter(AgeGroup == "3:Newest")
table(new_rcs$Design, new_rcs$Scale)

#subsetting a 3-way table
table(rcs$Design, rcs$Scale, rcs$AgeGroup, useNA = 'always')[,,"2:Recent"]
```

Lets use `group_by`and `summarize()` to create a two-way contingency table, then use `pivot_wider()` to mimic the output of `table()`

```{r}
rcs |>
  drop_na(Design, Type) |>
  group_by(Design, Type) |>
  summarize(count = n()) |>
  pivot_wider(names_from = Design, values_from = count)
```


Now we'll create some bar charts using `ggplot()`

```{r}
#stacked bar chart
ggplot(data = rcs |> drop_na(Design, Census.Region), aes(x = Census.Region, fill = Design)) + 
  geom_bar(position = "stack") + # set position to stack for stacked bar plot
  labs(title = 'Count of Roller Coaster Designs in Different Regions',
       x = 'Census Region',
       y = 'Count') +
  theme_minimal()

#side-by-side bar chart
ggplot(data = rcs |> drop_na(Design, Census.Region), aes(x = Census.Region, fill = Design)) + 
  geom_bar(position = "dodge") + # set position to dodge for side-by-side plot
  labs(title = 'Count of Roller Coaster Designs in Different Regions',
       x = 'Census Region',
       y = 'Count') +
  theme_minimal()
```

Seems like sit-down roller coasters are by far the most popular

## Numeric variables (and across groups)

Lets take a closer look at two of our numeric variables. I'll choose `TopSpeed` and `MaxHeight`.

First we'll examine the center and spread of these variables

```{r}
mean(rcs$TopSpeed, na.rm = TRUE)  # make sure to remove the NA values
sd(rcs$TopSpeed, na.rm = TRUE)

mean(rcs$MaxHeight, na.rm = TRUE)
sd(rcs$MaxHeight, na.rm = TRUE)
```

Now subset the data to include only inverted `Design` types. Then find their mean and median again

```{r}
inverted_design_rcs <- rcs |>
  filter(Design == "Inverted")

mean(inverted_design_rcs$TopSpeed, na.rm = TRUE)
sd(inverted_design_rcs$TopSpeed, na.rm = TRUE)

mean(inverted_design_rcs$MaxHeight, na.rm = TRUE)
sd(inverted_design_rcs$MaxHeight, na.rm = TRUE)
```

Seems like the inverted roller coasters have a higher mean `TopSpeed` and `MaxHeight`

Lets find measures of center and spread across a single grouping variable for two of these variables

```{r}
grouped1_rcs <- rcs |>
  select(Type, TopSpeed, MaxHeight) |>
  group_by(Type) |> # group by one variable
  drop_na() |> 
  summarize(across(everything(), # find the mean and sd across all variables
                   list(mean = mean, sd = sd)))

grouped1_rcs
```

Seems like wooden roller coasters have a higher `TopSpeed` on average compared to steel roller coasters. Their average `MaxHeight` is very similar, although the wooden coaster's `MaxHeight` is much less variable then the steel coaster's `MaxHeight`.

Now lets find measures of center and spread across two grouping variables for two of these variables/

```{r}
grouped2_rcs <- rcs |>
  select(Type, Design, TopSpeed, MaxHeight) |>
  group_by(Type, Design) |> # group by two variables
  drop_na() |> 
  summarize(across(everything(), 
                   list(mean = mean, sd = sd)))

grouped2_rcs
```

The steel type, wing design roller coasters seem like they're pretty scary. 

Now lets create a correlation matrix between all the numeric variables.

```{r}
rcs_numeric <- rcs |>
  select(where(is.numeric)) |> # select only the numeric variables from the rcs dataset
  drop_na() # drop missing values

cor(rcs_numeric)
```

The highest correlation I see is between `Drop` and `MaxHeight`. The lowest I see is between `Duration` and `Year.Opened`.

Lets create some graphs  that can compare the distributions across the groups on the same plot.

```{r}
# Histograms
ggplot(rcs |> select(Scale, TopSpeed) |> drop_na(), aes(x = TopSpeed, fill = Scale)) +
  geom_histogram() +
  labs(title = "Top Speed by Ride Type",
       x = "Top Speed (MpH)",
       y = "Frequency") +
  theme_bw() # Lets add a theme to our graphs

ggplot(rcs |> select(Scale, MaxHeight) |> drop_na(), aes(x = MaxHeight, fill = Scale)) +
  geom_histogram() +
  labs(title = "Max Height by Ride Type",
       x = "Height (ft.)",
       y = "Frequency") +
  theme_bw()
```


As expected, `TopSpeed` and `MaxHeight` seem to increase along with `Scale`. Kiddie and family rides have lower `TopSpeed` and `MaxHeight`

```{r}
# Boxplots
ggplot(rcs |> select(Scale, TopSpeed) |> drop_na(), aes(x = TopSpeed, fill = Scale)) +
  geom_boxplot() +
  labs(title = "Distribution of Top Speed by Scale of Ride",
       x = "Top Speed (MpH)",
       y = "Frequency") +
  theme_bw()

ggplot(rcs |> select(Scale, MaxHeight) |> drop_na(), aes(x = MaxHeight, fill = Scale)) +
  geom_boxplot() +
  labs(title = "Distribution of Max Height by Scale of Ride",
  x = "Height (ft.)",
  y = "Frequency") +
  theme_bw()
```

As expected, thrill and extreme rides have a higher `TopSpeed` and `Maxheight` on average, with kiddie and family rides having lower `TopSpeed` and `Maxheight`. Another thing I noticed is how little overlap there is in the spread of each `Scale` level. The middle 50% of the data within each `Scale` level almost never overlap for both `TopSpeed` and `Maxheight`. 

```{r}
# Kernel Density Plot
ggplot(rcs |> select(Scale, TopSpeed) |> drop_na(), aes(x = TopSpeed, fill = Scale)) +
  geom_density(alpha = 0.5) +
  labs(title = "Distribution of Top Speed by Scale of Ride",
       x = "Top Speed (MpH)",
       y = "Frequency") +
  theme_bw()

ggplot(rcs |> select(Scale, MaxHeight) |> drop_na(), aes(x = MaxHeight, fill = Scale)) +
  geom_density(alpha = 0.5) +
  labs(title = "Distribution of Max Height by Scale of Ride",
  x = "Height (ft.)",
  y = "Frequency") +
  theme_bw()
```

The density plots show a clear increase in `TopSpeed` and `MaxHeight` as ride `Scale` increases. Extreme rides have the highest speeds and heights with the widest distribution, while kiddie rides have the lowest speeds and heights with the smallest distribution. Extreme rides also exhibit a long right tail, indicating a few exceptionally fast coasters.


```{r}
# Scatter Plots
ggplot(rcs |> select(TopSpeed, MaxHeight, Scale) |> drop_na(), aes(x = MaxHeight, y = TopSpeed, 
                                                                   color = Scale)) +
  geom_point() +
  labs(title = "Correlation between Top Speed and Max Height",
       x = "Max Height (ft.)",
       y = "Top Speed (MpH)") +
  theme_bw()
```

This scatter plot tells us that `MaxHeight` increases along with `TopSpeed` with a noticably high correlation. We can also see that as `MaxHeight` and `TopSpeed` increase, ride `Scale` tends to increase as well. The frequency of extreme rides increase as `MaxHeight` and `TopSpeed` increase, while the frequency of kiddie and family rides decrease.

```{r}
ggplot(rcs |> select(Drop, Length, Type) |> drop_na(), aes(x = Drop, y = Length, 
                                                           color = Type)) +
  geom_point() +
  labs(title = "Correlation between Top Speed and Max Height",
       x = "Max Height (ft.)",
       y = "Top Speed (MpH)") +
  theme_bw()
```

The correlation between `Drop` and `Length` does not seem as strong. We can see that steel roller coasters are a bit more variable in `Drop`, but apart from that, there is not much this graph is telling us. 

```{r}
# Faceting the scatter plots with one categorical variable
ggplot(rcs |> select(TopSpeed, MaxHeight, Scale) |> drop_na(), aes(x = MaxHeight, y = TopSpeed)) +
  geom_point() +
  labs(title = "Top Speed by Max Height",
       x = "Max Height (Feet)",
       y = "Top Speed (MPH)") +
  facet_wrap(~Scale, nrow = 2) +
  theme_bw()

ggplot(rcs |> select(Drop, Length, Type) |> drop_na(), aes(x = Drop, y = Length)) +
  geom_point() +
  labs(title = "Correlation between Top Speed and Max Height",
       x = "Max Height (ft.)",
       y = "Top Speed (MpH)") +
  facet_wrap(~Type, nrow = 2) +
  theme_bw()
```



```{r}
# Faceting the scatter plots with two categorical variables
ggplot(rcs |> select(TopSpeed, MaxHeight, Scale, Type) |> drop_na(), aes(x = MaxHeight, y = TopSpeed)) +
  geom_point() +
  labs(title = "Top Speed by Max Height",
       x = "Max Height (Feet)",
       y = "Top Speed (MPH)") +
  facet_wrap(vars(Type,Scale)) +
  theme_bw()

ggplot(rcs |> select(Drop, Length, Type, Scale) |> drop_na(), aes(x = Drop, y = Length)) +
  geom_point() +
  labs(title = "Correlation between Top Speed and Max Height",
       x = "Max Height (ft.)",
       y = "Top Speed (MpH)") +
  facet_wrap(vars(Type,Scale)) +
  theme_bw()
```

