---
title: "ST 558 HW 4 - Chris's version"
format: html
---

## Chris's work goes here

# Task 1: Querying a News API
First, we're querying the news API:
```{r}
library(httr)
library(jsonlite)

# API call
news_return <- GET("https://newsapi.org/v2/everything?q=Iran&from=2026-05-13&language=en&pageSize=50&apiKey=d41f8e9cca0f4687b4865289b712e088")

str(news_return)
```
We've got the data, but most of this looks like metadata we aren't interested in. The articles are stored in the `content` object. We need to parse the raw data using the `jsonlite` package:

```{r}
suppressPackageStartupMessages(library(tidyverse))

parsed <- fromJSON(rawToChar(news_return$content))
articles <- parsed$articles |>
            as_tibble()

articles
```
Looks great! Now we'll write a function to query the API more easily:
```{r}
query_news <- function(search, start_date, api_key) {
  # build url
  url <- paste0("https://newsapi.org/v2/everything?q=",
                search,
                "&from=",
                start_date,
                "&language=en&pageSize=50&apiKey=",
                api_key)
  # call API
  api_return <- GET(url)
  
  # parse return and output as tibble
  parsed <- fromJSON(rawToChar(api_return$content))
  articles <- parsed$articles |>
              as_tibble()
  return(articles)
}

# 1) original API call
query_news("Iran", 2026-05-11, "d41f8e9cca0f4687b4865289b712e088")

# 2) articles about AI from this month
query_news("AI", 2026-06-01, "d41f8e9cca0f4687b4865289b712e088")
```
It works!

# Task 2: Summarizing Roller Coaster Data
Next, we're reading in data about roller coasters. Let's check out the dataset:
```{r}
rcs_raw <- read.csv("https://www4.stat.ncsu.edu/online/datasets/635USrollercoasters.csv")

# remove unnecessary columns
rcs <- rcs_raw |>
       select(-c(Boundary, Latitude, Longitude))

# look at types
str(rcs)
```
Looking at these variables and their types, I think we could do some cleaning. Specifically, I see the following:

- `Census.Region`, `AgeGroup`, `Scale`, `type`, `design`, and `inversions` should be factors
- The "number of inversions" and the "inversions." variable need to be renamed

```{r}
rcs <- rcs_raw |>
       select(-c(Boundary, Latitude, Longitude)) |>
       rename(Inversions =`Inversions.`,
              Num.Inversions= `X..of.Inversions`) |>
       mutate(across(c(Census.Region, AgeGroup, Scale, Type, Design, Inversions),
                          ~ as.factor(.))) |>
       as_tibble()

rcs
```
Now that our data is clean, we should see if there are any missing values to note. I'm finding the proportion of missing values so that we can get a better idea of the missingness:
```{r}
round(colSums(is.na(rcs))/nrow(rcs), 3) |>
  sort(decreasing = TRUE)
```
Checking out the output, we can see that the `Drop`, `TopSpeed`, `Duration`, and `Length` variables have the greatest share of missing values. Respectively, they have missingness 20%, %12, %10, and %5. 

## Exploring Categorical Variables:

### Contingency Tables with `table()`
```{r}
table(rcs$AgeGroup)
table(rcs$AgeGroup, rcs$Scale)
# interpreting this one Vvv
table(rcs$AgeGroup, rcs$Scale, rcs$Type)
```
The value "16" in the second level of the two way table implies that there are 16 older wooden thrill rides in the dataset.

### Conditional Tables
```{r}
# way one - subsetting the data
wood_rcs <- filter(rcs, Type == "Wood")
table(wood_rcs$AgeGroup, wood_rcs$Scale)

# way two - subsetting a three way table
table(rcs$AgeGroup, rcs$Scale, rcs$Type)[,,2]
```
### Contingency Tables with `group_by()` and `summarize()`
```{r}
dplyr_table <- rcs |>
               select(AgeGroup, Scale, Type) |>
               group_by(Type, AgeGroup, Scale) |>
               dplyr::summarize(count = n()) |>
               pivot_wider(names_from = Scale, values_from = count)


dplyr_table
```


### Stacked and Side-by-side Bar Graphs
Stacked plots are the default option, so we just use `geom_bar()`
```{r}
library(ggplot2)

stacked_bar <- ggplot(rcs, aes(x = Scale, fill = Type)) +
               geom_bar() +
               labs(title = "Count By Roller Coaster Scale",
                    subtitle = "Stacked Bars",
                    y = "Frequency") +
               theme_minimal()

stacked_bar
```
To get the bars to dodge, we just have to specify the `position` argument:
```{r}
library(ggplot2)

sbs_bar <- ggplot(rcs, aes(x = Scale, fill = Type)) +
           geom_bar(position = "dodge") +
           labs(title = "Count By Roller Coaster Scale",
                subtitle = "Side-by-side Bars",
                y = "Frequency") +
           theme_minimal()

sbs_bar
```

## Exploring Numeric Variables
### Center + Spread for two variables
Now, we're looking at the mean and standard deviation for `TopSpeed` and `MaxHeight`:
```{r}
cntr_sprd_tbl <- rcs |>
                 select(TopSpeed, MaxHeight) |>
                 drop_na() |> # drop missing due to high missingness for TopSpeed
                 summarize(across(everything(),
                                  list(mean = mean, sd = sd)))

cntr_sprd_tbl
```
Now subset to only include steel coasters:
```{r}
cntr_sprd_tbl_steel <- rcs |>
                       filter(Type == "Steel") |>
                       select(TopSpeed, MaxHeight) |>
                       drop_na() |> # drop missing due to high missingness for TopSpeed
                       summarize(across(everything(),
                                        list(mean = mean, sd = sd)))

cntr_sprd_tbl_steel
```
### Center + Spread for two variables + Grouping
```{r}
cntr_sprd_1grp <- rcs |>
                  select(Type, TopSpeed, MaxHeight) |>
                  group_by(Type) |>
                  drop_na() |> # drop missing due to high missingness for TopSpeed
                  summarize(across(everything(),
                                  list(mean = mean, sd = sd)))

cntr_sprd_1grp
```

```{r}
cntr_sprd_2grp <- rcs |>
                  select(Type, Scale, TopSpeed, MaxHeight) |>
                  group_by(Type, Scale) |>
                  drop_na() |> # drop missing due to high missingness for TopSpeed
                  summarize(across(everything(),
                                  list(mean = mean, sd = sd)))

cntr_sprd_2grp
```
As we saw previously, there isn't enough data to estimate the height and speed of the wooden kiddie coasters.

### Correlation Matrix
```{r}
rcs_numeric <- select(rcs, where(is.numeric)) |>
               drop_na()


cor(rcs_numeric)
```

### Histogram
```{r}
top_spd_hist <- ggplot(rcs |> select(Type, TopSpeed) |> drop_na(), aes(x = TopSpeed, fill = Type)) +
                geom_histogram() +
                labs(title = "Top Speed by Ride Type",
                     x = "Top Speed (MpH)",
                     y = "Frequency")

max_hgt_hist <- ggplot(rcs |> select(Type, MaxHeight) |> drop_na(), aes(x = MaxHeight, fill = Type)) +
                geom_histogram() +
                labs(title = "Maximum Height by Ride Type",
                     x = "Height (ft.)",
                     y = "Frequency")


top_spd_hist
max_hgt_hist
```

### Density Plots
```{r}
top_spd_density <- ggplot(rcs |> select(Type, TopSpeed) |> drop_na(), aes(x = TopSpeed, fill = Type)) +
                   geom_density(alpha = 0.5) +
                   labs(title = "Distribution of Top Speed by Ride Type",
                        x = "Top Speed (MpH)",
                        y = "Frequency")

max_hgt_density <- ggplot(rcs |> select(Type, MaxHeight) |> drop_na(), aes(x = MaxHeight, fill = Type)) +
                   geom_density(alpha = 0.5) +
                   labs(title = "Distribution of Maximum Height by Ride Type",
                        x = "Height (ft.)",
                        y = "Frequency")


top_spd_density
max_hgt_density
```
### Box Plots
```{r}
top_spd_box <- ggplot(rcs |> select(Type, TopSpeed) |> drop_na(), aes(x = TopSpeed, fill = Type)) +
               geom_boxplot() +
               labs(title = "Distribution of Top Speed by Ride Type",
                        x = "Top Speed (MpH)",
                        y = "Frequency")

max_hgt_box <- ggplot(rcs |> select(Type, MaxHeight) |> drop_na(), aes(x = MaxHeight, fill = Type)) +
               geom_boxplot() +
               labs(title = "Distribution of Maximum Height by Ride Type",
                        x = "Height (ft.)",
                        y = "Frequency")


top_spd_box
max_hgt_box
```
### Scatterplots
```{r}
topspd_maxhgt <- ggplot(rcs |> drop_na(), aes(x = MaxHeight, y = TopSpeed, color = Type)) +
                 geom_point() +
                 labs(title = "Top Speed by Max Height",
                      x = "Max Height (Feet)",
                      y = "Top Speed (MPH)") +
                 theme_minimal()


topspd_maxhgt
```

### Faceting:
```{r}
topspd_maxhgt_1fct <- ggplot(rcs |> drop_na(), aes(x = MaxHeight, y = TopSpeed)) +
                      geom_point() +
                      labs(title = "Top Speed by Max Height",
                               x = "Max Height (Feet)",
                               y = "Top Speed (MPH)") +
                      facet_wrap(~Type, nrow = 2) +
                      theme_minimal()

topspd_maxhgt_1fct
```

```{r}
topspd_maxhgt_2fct <- ggplot(rcs |> drop_na(), aes(x = MaxHeight, y = TopSpeed)) +
                      geom_point() +
                      labs(title = "Top Speed by Max Height",
                               x = "Max Height (Feet)",
                               y = "Top Speed (MPH)") +
                      facet_wrap(vars(Type, Scale)) +
                      theme_minimal()

topspd_maxhgt_2fct
```










